---
title: "BIOST 544 Homework 1"
author: "Hantong Hu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = F)
```

# Analysis Questions

1. The point estimate and 90% interval of the probability a patient on TFD725+docetaxel will survive past 400 days in each age category is presented below.

```{r q1, warning=FALSE, message=FALSE, echo =FALSE}
## Q1
nsclc <- read.table("nsclc-modified.txt", header = TRUE)

nsclc.50 <- nsclc %>% mutate(age.cat = case_when(50<=age & age<55 ~ "50+",
                                                 55<=age & age<60 ~ "55+",
                                                 60<=age & age<65 ~ "60+",
                                                 65<=age & age<70 ~ "65+", 
                                                 70<=age ~ "70+")) %>%
  filter(!is.na(age.cat) & tx==1)

# Point estimation
prop.table <- nsclc.50 %>% group_by(age.cat) %>% summarise(n=n(), prop = mean(survival.past.400))

# Interval estimation
calc_sample_dist <- function(nsamp, subject.num, pi){
  sample_means <- rbinom(nsamp, subject.num, pi)/subject.num
  return(sample_means)
}

candidate_pi_101 <- seq(from = 0, to = 1, length.out = 101)
CI.table <- data.frame(prop.table, lower=NA, higher=NA)

set.seed(1)
for (i in c(1:nrow(CI.table))){
  percentiles_101 <- c()
  for(pi in candidate_pi_101){
    samp_dist <- calc_sample_dist(10000, CI.table[i,2],pi)
    percentile <- mean(samp_dist <= CI.table[i,3])
    percentiles_101 <- c(percentiles_101,percentile)
  }
  
  consistent_pi <- candidate_pi_101[(percentiles_101 >= 0.05) & (percentiles_101 <= 0.95)]
  CI.table[i,4] <- min(consistent_pi)
  CI.table[i,5] <- max(consistent_pi)
}
colnames(CI.table) <- c("Age category", "Count", "Point estimate", "90% CI lower", "90% CI higher")
knitr::kable(CI.table, digits = 2, 
             caption = "Point estimation and 90% interval for patients on TFD725+docetaxel")
```

2. We first want to list the table of number of patient and estimated probability of survival past 400 days for both treatment arms and each age category, and we get the following table. The table also includes the observed difference of proportions and overall proportion. 

```{r q2a, warning=FALSE, message=FALSE, echo =F}
## Q2
nsclc.50.control <- nsclc %>% mutate(age.cat = case_when(50<=age & age<55 ~ "50+",
                                                 55<=age & age<60 ~ "55+",
                                                 60<=age & age<65 ~ "60+",
                                                 65<=age & age<70 ~ "65+", 
                                                 70<=age ~ "70+")) %>%
  filter(!is.na(age.cat) & tx==0)
nsclc.50.all <- rbind(nsclc.50, nsclc.50.control)

# Count and prop and prop diff for two tx for each age category
colnames(prop.table) <- c("age.cat", "trt.count", "trt.prop")
all.prop.table <- cbind(prop.table,
                        (nsclc.50.control %>% group_by(age.cat) %>% 
                          summarise(ctr.count = n(), ctr.prop = mean(survival.past.400)))[,-1] )
all.prop.table$prop.diff <- all.prop.table$trt.prop - all.prop.table$ctr.prop
all.prop.table <- cbind(all.prop.table, (nsclc.50.all %>% group_by(age.cat) %>%
                                  summarise(prop.overall=mean(survival.past.400)))[,-1])

all.table <- all.prop.table
colnames(all.table) <- c("Age category", "TFD725 count", "TFD725 prop", 
                         "Control count", "Control prop", "Prop diff", "Prop overall")
knitr::kable(all.table, digits = 2, caption = "Proportion for 2 arms, overall and difference")
```

Next, we want to evaluate whether TFD725 is more effective than control. If TFD725 has the same effect as the control, we would expect the TFD725 proportion and Control proportion both equal to the overall proportion. We can look at the histogram of simulated proportion differences. The red lines show the observed difference of proportions for 2 arms. Below the histograms, we also display where the observed difference lie on the histogram (percent to the left) as a table.

```{r q2b, warning=FALSE, message=FALSE, echo =F}
simulate.trial <- function(pi.treat, pi.control, n.treat, n.control){
    patients.treat <- rbinom(1,n.treat,pi.treat)
    patients.control <- rbinom(1,n.control,pi.control)

    prop.diff <- patients.treat/n.treat - patients.control/n.control

    return(prop.diff)
}

ntrial <- 10000
sim.prop.diff <- data.frame(age.cat=character(), simulated.prop.diffs=double())
pct <- c()
set.seed(1)

for (i in c(1:nrow(all.prop.table))){
  simulated.prop.diffs <- replicate(ntrial,
                                  simulate.trial(all.prop.table[i,7], all.prop.table[i,7], 
                                                 all.prop.table[i,2], all.prop.table[i,4]))
  sim.prop.diff <- rbind(sim.prop.diff, data.frame(age.cat=all.prop.table[i,1],
                                                      simulated.prop.diffs))
  
  
  # p <- ggplot(data = as.data.frame(simulated.prop.diffs), aes(simulated.prop.diffs)) + 
  #   geom_histogram(alpha=0.4, colour="black", fill="blue") +
  #   ggtitle(paste("Age group: ", all.prop.table[i,1]))
  # plot.ls <- c(plot.ls, p)
  
  
  pct <- c(pct, mean(simulated.prop.diffs <= all.prop.table[i,6]))
}

sim.prop.diff <- merge(sim.prop.diff, all.prop.table[,c(1,6)])
p <- ggplot(sim.prop.diff, aes(x=simulated.prop.diffs)) +
    geom_histogram(aes(y=..count..), colour='black', fill='skyblue') + 
    geom_density() +
    facet_wrap(~age.cat, scales="free")
p + geom_vline(data=sim.prop.diff, aes(xintercept=prop.diff), colour="red") + 
  ggtitle("Histogram of simulated proportion differences")

percent <- data.frame(age.cat=all.prop.table$age.cat, pct=pct*100)
colnames(percent) <- c("Age category", "Percent to the left")
knitr::kable(percent, caption = "Place of observed difference on histograms")
```

By the above table and figures, we conclude that for age groups 50+ and 65+, TFD725 is more effective than control, while the other age groups don't have significant differences. To evaluate if the treatment effect appears to substantively and/or systematically differ across age, we can fit a linear regression of observation time on age and treatment. From the table below, we conclude that for two patients with one year difference in age in the control group, the elder patient would be expected to live 0.004 days longer, while for two patients with one year difference in age in the treatment group, the elder patient would be expected to live 0.678 days shorter. We have no evidence that the treatment
effect appears to substantively and/or systematically differ across age (P>0.05).

```{r q2c, warning=FALSE, message=FALSE, echo =F}
mod <- lm(obstime~age*tx, data = nsclc)
coef <- summary(mod)$coefficients
knitr::kable(coef[,c(1,4)], digits = 3, caption="Point estimation and P value for linear regression of observation time on age and treatment")
```

\pagebreak

# Code
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```