---
title: "BIOST 544 Homework 3"
author: "Hantong Hu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = F)
```

## Data Preparation

Before investigating the relationship between gene-expression values in the tumor and the existence and extent of necrotic tissue, we would first want to prepare the proper data set for this problem. After joining the clinical data and gene expression data by "centerid" and "patid", we want to have an overview of the necrotic cell percent for all patients. The plot below shows the distribution of necrotic cell percent for patients.

```{r prep, warning=FALSE, message=FALSE, echo =FALSE}
## Data preparation
setwd("C:/Users/second/Desktop/BIOST 544/Homework/HW3")

library(data.table)
NOAH.clinical <- read.csv("clinical_data.csv", header = T)[,-1]
NOAH.clinical.keep <- NOAH.clinical[,c(1,2,11)]
NOAH.anno <- read.csv("annotation.csv", header = T)
NOAH.expression <- fread("expression_data_probeID.csv", header=T, sep=',')[,-1]

if(typeof(NOAH.expression$patid) != typeof(NOAH.clinical.keep$patid)){
  warning("patid types dont match! Inverting both to numeric")
  NOAH.expression$centerid <- as.numeric(NOAH.expression$centerid)
  NOAH.expression$patid <- as.numeric(NOAH.expression$patid)
}

# NOAH.exp.keep <- NOAH.expression[,c(1:102)]
# head(NOAH.exp.keep[,c(1:10)])
# NOAH <- inner_join(NOAH.clinical.keep, NOAH.exp.keep, by=c("centerid","patid"))

NOAH <- inner_join(NOAH.clinical.keep, NOAH.expression, by=c("centerid","patid"))

ggplot(data = NOAH[,c(1:3)], aes(x=necrotic_cells.pct, y=..density..)) + 
  geom_density() +
  geom_histogram(fill=2, alpha=0.2) +
  ggtitle("Distribution of necrotic cell percent for patients")
```

## Screening-based

We first look at the spearman correlation between each gene and necrotic cell percent for each patient, and, for convenience, we choose to include at most 50 genes with the highest absolute value of correlation for further investigation. Then we want to know how many of the top genes can give the best prediction of necrotic cell percent by fitting 50 linear models with the top n (n from 1 to 50) correlated genes and see which one has the smallest MSE. We want to train the model by splitting the original data set with train and test data.

```{r screening, warning=FALSE, message=FALSE, echo =F}
## Screening
# Split into train and test data
set.seed(1)
NOAH.split <- NOAH %>% mutate(train.test= sample(c(0,1), nrow(NOAH), replace = TRUE))
NOAH.train <- NOAH.split[which(NOAH.split$train.test==1), c(3:(ncol(NOAH.split)-1))]
NOAH.test <- NOAH.split[which(NOAH.split$train.test==0), c(3:(ncol(NOAH.split)-1))]

# Calculate correlation between necrotic and each gene from training data
gene.cor <- data.frame(cor=t(cor(NOAH.train$necrotic_cells.pct, NOAH.train[,-1], 
                                 method = "spearman")))

# Find the top 50 genes with highest absolute value and record gene names
gene.cor$abs.cor <- abs(gene.cor$cor)
gene.list <- gene.cor %>% arrange(desc(abs.cor)) %>% head(50) %>% rownames


MSEs <- rep(NA,times=50)
for(i in c(1:50)){
  lm.train <- lm(necrotic_cells.pct~., data=NOAH.train[,c("necrotic_cells.pct",gene.list[1:i])])
  preds <- predict(lm.train, NOAH.test[,c("necrotic_cells.pct",gene.list[1:i])])
  MSEs[i] <- mean((NOAH.test$necrotic_cells.pct - preds)^2)
}

```

After this process, we find that the top `r which.min(MSEs)` genes give the smallest MSE. The genes (with annotations) are listed as below:

```{r screening report, warning=FALSE, message=FALSE, echo =F}
top.gene <- data.frame(probset.ids=gene.list[1:which.min(MSEs)])
top.gene.name <- inner_join(top.gene, NOAH.anno, by=c("probset.ids"))
knitr::kable(top.gene.name, caption = "Top genes selected for the best fit")
```

## Prediction-based

For this method, we will use the cv.glmnet function to build a model to predict necrotic cell percent by using all genes. Below is the plot for this model. However, the results show that the optimal model has 0 features, in another word, only the intercept has a non-zero value. Thus we can say the prediction-based model is not viable in this question.

```{r prediction, warning=FALSE, message=FALSE, echo = F}
## Prediction
library(glmnet)
fit.cv <- cv.glmnet(as.matrix(NOAH[,-c(1:3)]), NOAH$necrotic_cells.pct)
plot(fit.cv)

# sum(coef(fit.cv, s=fit.cv$lambda.min)!=0)
```


\pagebreak

# Code
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```