---
title: "BIOST 544 Homework 4"
author: "Hantong Hu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = F)
```

## Data Preperation

To answer the question about the relationship between smoking and bone health, we can approach it by comparing the BMD in patients who have vs. have not smoked. There are some confounders we think may be of interest in this problem, so we will first address the confounders, and then specify our exposure and outcome. The confounders we will consider in this problems are: age of the patients, alcohol in the past 24 hrs, number of person smoking in the house.

For the exposure, we will use the variable SMOKERE0, i.e. regular smoker, to define smokers, and patients with missing smoking status value will be omitted. BMD in this data set is presented in two variables: Total Hip BMD (HPBMDT0) and Total Spine BMD (SPBMDT0). All patients with any missing entries will be omitted from the data for convenience, which leads to a total of 2233 patients.

```{r data prep, warning=FALSE, message=FALSE, echo =FALSE, fig.height=3, fig.width=3, fig.align='center'}
## Data prep
library(foreign)
swan <- read.dta("28762-0001-Data.dta")

# summary(swan$SMOKERE0)
# summary(swan$HPBMDT0)
# summary(swan$SPBMDT0)

swan.use <- swan %>% select(SMOKERE0, HPBMDT0, SPBMDT0, AGE0, ALCHL240, HHMEMSM0)
swan.use$smoker <- ifelse(swan.use$SMOKERE0 %in% c("Missing","Do not know","NA's"),
                          NA,ifelse(swan.use$SMOKERE0=="Yes",1,0))
swan.use$alcohol <- ifelse(swan.use$ALCHL240 %in% c("Missing","N/A"),
                           NA,ifelse(swan.use$ALCHL240=="Yes",1,0))
swan.use$housemem <- ifelse(swan.use$HHMEMSM0<0,NA,swan.use$HHMEMSM0)

swan.complete <- swan.use[,c(-1,-5,-6)] %>% na.omit()
```

Since we have hip BMD and spine BMD, we will explore these two variables separately. First we want to look at the distribution of these two variables among smokers and non-smokers by plotting a boxplot.

```{r boxplot, warning=FALSE, message=FALSE, echo =FALSE, fig.height=3, fig.width=3}
## Boxplot for BMD
library(ggplot2)
ggplot(data = swan.complete, aes(x=as.factor(smoker), y=HPBMDT0)) +
  geom_boxplot() + xlab("Smoker")+ylab("Total Hip BMD")
ggplot(data = swan.complete, aes(x=as.factor(smoker), y=SPBMDT0)) +
  geom_boxplot()+ xlab("Smoker")+ylab("Total Spine BMD")
```

From the above boxplots, we do not see much of a difference (both for hip and spine) in BMD between smokers and non-smokers. Thus we want to use permutation to gain a more solid conclusion.

## Permutation

```{r perm, warning=FALSE, message=FALSE, echo =F, results='hide'}
## Permutation w/o confounders
obs.diff.hp <- with(swan.complete, mean(HPBMDT0[smoker==1]) - mean(HPBMDT0[smoker==0]))
obs.diff.sp <- with(swan.complete, mean(SPBMDT0[smoker==1]) - mean(SPBMDT0[smoker==0]))

one.perm.hp <- function(data){
  perm <- sample(1:nrow(data), replace = FALSE)
  perm.data <- data
  perm.data$smoker = data$smoker[perm]
  
  perm.diff <- with(perm.data, mean(HPBMDT0[smoker==1]) - mean(HPBMDT0[smoker==0]))
  return(perm.diff)
}

one.perm.sp <- function(data){
  perm <- sample(1:nrow(data), replace = FALSE)
  perm.data <- data
  perm.data$smoker = data$smoker[perm]
  
  perm.diff <- with(perm.data, mean(SPBMDT0[smoker==1]) - mean(SPBMDT0[smoker==0]))
  return(perm.diff)
}
```

The observed difference for smokers and non-smokers for hip BMD is `r round(obs.diff.hp,3)`, and for spine BMD is `r round(obs.diff.sp,3)`. To evaluate if smokers and non-smokers have significant difference in BMD, we will perform permutation tests on both hip and spine BMD. Below are mean differences generated by permutation compared to the original difference in the data set.

```{r perm test, warning=FALSE, message=FALSE, echo =F, fig.width=3, fig.height=3}
set.seed(1)
permuted.stats <- data.frame(hp.perm=replicate(1e3, one.perm.hp(swan.complete)), 
                             sp.perm=replicate(1e3, one.perm.sp(swan.complete)))

ggplot(permuted.stats, aes(x=hp.perm, y=..density..)) +
    geom_density()+xlab("Hip BMD mean diff")+
    geom_vline(xintercept=obs.diff.hp, colour = "red")

ggplot(permuted.stats, aes(x=sp.perm, y=..density..)) +
    geom_density()+xlab("Spine BMD mean diff")+
    geom_vline(xintercept=obs.diff.sp, colour = "red")
```

```{r perm p value, warning=FALSE, message=FALSE, echo =F, results='hide'}
p.hp <- mean(permuted.stats$hp.perm>obs.diff.hp)
p.sp <- mean(permuted.stats$sp.perm>obs.diff.sp)
```

The p-value for Hip BMD is `r p.hp` and Spine BMD is `r p.sp`, so we have strong evidence to reject the null hypothesis that there is no relationship between smoking and BMD, if not adjusting for confounders.

## Confounding

Next we want to see if the above results still holds if we take confounding variables into account. As stated in the first section, age, alcohol intake in the past 24 hrs, and number of members who smoke in house, will be taken into account as confounding variables. To address the confounders, we use the propensity scores/inverse probability weighting. Below is the density plot for the propsensity score calculated from logistic regression.

```{r Propensity, warning=FALSE, message=FALSE, echo = F, fig.width=5, fig.height=3, fig.align='center'}
## Propensity
propen.model <- glm(smoker ~ AGE0 + alcohol + housemem, family=binomial, data=swan.complete)

propensities <- predict(propen.model, data=swan.complete, type="response")

ggplot(data.frame(propensities=propensities, smoker=swan.complete$smoker), 
       aes(x = propensities, y = ..density.., color = as.factor(smoker))) + geom_density()

# range(propensities)
trunc.propen <- propensities %>% pmin(0.95) %>% pmax(0.05)
# range(trunc.propen)
```

Our propensity scores range from `r round(range(propensities),2)`, where the tail is very close to 1, so we will truncate the 5% both on the head and tail to keep the propensity score away from 0 and 1. The truncated range is from `r round(range(trunc.propen),2)`. Next we will perform the inverse weight. The ratio of those enrollment probabilities is shown as below.

```{r ipw, warning=FALSE, message=FALSE, echo =F, fig.height=3, fig.width=4, fig.align='center'}
## Inverse probability weighting
npat <- nrow(swan.complete)
weights <- rep(0, npat)

## for patients who smoke:
representative.propen <- sum(swan.complete$smoker) / npat
actual.propen <- trunc.propen

smoke.ind <- which(swan.complete$smoker == 1)
weights[smoke.ind] <- representative.propen/actual.propen[smoke.ind]
weights[-smoke.ind]<- (1 - representative.propen)/(1 - actual.propen[-smoke.ind])

ggplot(data.frame(weights = weights), aes(x=weights, y =..density..)) + 
  geom_histogram()

diff.est.hp <- with(swan.complete, mean((weights*HPBMDT0)[smoke.ind])) - 
  with(swan.complete, mean((weights*HPBMDT0)[-smoke.ind]))
diff.est.sp <- with(swan.complete, mean((weights*SPBMDT0)[smoke.ind])) - 
  with(swan.complete, mean((weights*SPBMDT0)[-smoke.ind]))
```

The mean difference estimated for smokers and non-smokers for weighted BMD is, for hip `r round(diff.est.hp ,3)`, and for spine `r round(diff.est.sp ,3)`. As a reminder, the observed difference for smokers and non-smokers for hip BMD is `r round(obs.diff.hp,3)`, and for spine BMD is `r round(obs.diff.sp,3)`. Both of the mean difference estimated are smaller than the observed difference. We want to see if this is due to random chance. After we rerandomized sampling distribution, we can compare our estimated mean difference to the histogram we get under re-randomization.

```{r re-rand, warning=FALSE, message=FALSE, echo =F, fig.height=3, fig.width=3}
## Rerandomization
do.one.propen <- function(outcome, propen){
  n <- length(outcome)
  label <- rbinom(n,1,propen)

  weights <- rep(0,n)  
  representative <- mean(label)
  actual <- propen
  ind.t <- which(label == 1)
  weights[ind.t] <- (representative/actual)[ind.t]
  weights[-ind.t] <- ((1-representative)/(1-actual))[-ind.t]
  
  return(mean((weights*outcome)[ind.t]) - mean((weights*outcome)[-ind.t]))
}

set.seed(1)
rerandomized.diffs.hp <- 
  replicate(1e3, do.one.propen(swan.complete$HPBMDT0, trunc.propen))
rerandomized.diffs.sp <- 
  replicate(1e3, do.one.propen(swan.complete$SPBMDT0, trunc.propen))

ggplot(data.frame(diffs = rerandomized.diffs.hp), aes(x = diffs, y = ..density..)) +
  geom_density() +xlab("Hip BMD mean diff")+
  geom_vline(xintercept = diff.est.hp, color = "red")
ggplot(data.frame(diffs = rerandomized.diffs.sp), aes(x = diffs, y = ..density..)) +
  geom_density() +xlab("Spine BMD mean diff")+
  geom_vline(xintercept = diff.est.sp, color = "red")
```

```{r rerand p value, warning=FALSE, message=FALSE, echo =F, results='hide'}
p.hp.rr <- mean(rerandomized.diffs.hp>diff.est.hp)
p.sp.rr <- mean(rerandomized.diffs.sp>diff.est.sp)
```

In the rerandomization process, we see that the p-value for Hip BMD is `r p.hp.rr` and Spine BMD is `r p.sp.rr`, so we do not have enough evidence to reject the null hypothesis that there is no relationship between smoking and BMD, if adjusting for confounders. Remember the p-value for permutation previously is `r p.hp` for hip BMD and `r p.sp` for Spine BMD. Thus, there is a huge difference if addressing for confounder.

## Standardization

We will use the standardization process, or fit a linear regression model including all confounders we listed above, to both of the BMDs and see the effect of smoking. The result of the three methods: observed mean difference, estimated mean difference using IPW, and standardization, are listed in the table below.

```{r standardization, warning=FALSE, message=FALSE, echo =F}
outcome.regression.hp <- lm(HPBMDT0 ~ smoker + AGE0 + alcohol + housemem, 
                            family=binomial, data=swan.complete)
outcome.regression.sp <- lm(SPBMDT0 ~ smoker + AGE0 + alcohol + housemem, 
                            family=binomial, data=swan.complete)

swan.smoker <- swan.complete %>% mutate(smoker = 1)
swan.nonsmoker <- swan.complete %>% mutate(smoking = 0)

standardized.est.hp <- mean(predict(outcome.regression.hp, swan.smoker, type = "response") -
                              predict(outcome.regression.hp, swan.nonsmoker, type = "response"))
standardized.est.sp <- mean(predict(outcome.regression.sp, swan.smoker, type = "response") -
                              predict(outcome.regression.sp, swan.nonsmoker, type = "response"))

summ <- data.frame(HPBMD=c(obs.diff.hp, diff.est.hp, standardized.est.hp), 
                   SPBMD=c(obs.diff.sp, diff.est.sp, standardized.est.sp))
rownames(summ) <- c("Observed", "IPW", "Standardization")
knitr::kable(summ, digits = 3)
```


\pagebreak

# Code
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```